---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kioto

<!-- badges: start -->
[![R-CMD-check](https://github.com/shikokuchuo/kioto/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/shikokuchuo/kioto/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```
  ________
 /\ m     \
/  \ o     \
\  /  v    /
 \/_______/
```

Parallel computation and data exchange between R processes on the same
machine: lock-free channels and work-stealing task pools over POSIX
shared memory (Linux, macOS) or Win32 file mappings (Windows).
The hot path stays entirely in user space: single-producer
single-consumer rings with batched publication, hybrid spin-then-park
waiting, and event-driven peer-death detection.

kioto is a complement to [mirai](https://mirai.r-lib.org) for workloads
measurably bound by per-message socket overhead. Payload contents
interoperate transparently with [mori](https://github.com/r-lib/mori)
shared objects.

## Installation

Install the development version from GitHub:

``` r
pak::pak("shikokuchuo/kioto")
```

kioto requires 64-bit R, and on Linux a kernel >= 5.3.

## Quick start

The essential function is `kio_channel()`: it creates a bidirectional
shared-memory channel — one lock-free ring per direction — and spawns a
child R process connected to its other end. The child evaluates a quoted
expression with `ch` bound to its side of the channel; all data crosses the
rings, not the process boundary.

```{r channel}
library(kioto)

ch <- kio_channel(quote(
  repeat {
    x <- kio_recv(ch, timeout = 30)
    if (inherits(x, "kio_condition")) break
    kio_send(ch, x * 2)
    kio_flush(ch)
  }
))

kio_send(ch, 21)
kio_flush(ch)
kio_recv(ch, timeout = 5)
```

`kio_send()` stages a message; `kio_flush()` publishes staged messages to
the peer — flush after every send for minimum latency, or every N sends for
throughput. Terminal states (ring full, timeout, orderly close, peer death)
are returned as class-tagged sentinels rather than errors, so the hot loop
stays branch-cheap; `kio_send_batch()` and `kio_recv_batch()` amortize the
R call boundary itself.

`kio_close()` performs an orderly shutdown, rendezvousing with the peer
before releasing shared resources:

```{r close}
kio_close(ch)
```

## Task pool

Built on the same transport, `kio_pool()` spawns a pool of worker processes
with work-stealing deques and no dispatcher in the loop: submission is a
shared-memory ring write plus at most one directed wake.

```{r pool}
p <- kio_pool(n_workers = 4L)

t <- kio_submit(p, sum(x) + y, x = runif(10), y = 100)
kio_collect(t)
```

Task expressions see their evaluating worker's own handle as `pool`, so
tasks can fan out nested subtasks. A nested submit pushes straight onto the
worker's own work-stealing deque — no ring, no wake — and a worker blocked
collecting a nested result executes other work instead of sleeping, so
divide-and-conquer runs at fork/join cost and never deadlocks the pool:

```{r nested}
t <- kio_submit(
  p,
  {
    subtasks <- lapply(parts, function(part) kio_submit(pool, sum(x), x = part))
    do.call(sum, lapply(subtasks, kio_collect))
  },
  parts = split(1:1000, rep(1:4, each = 250))
)
kio_collect(t)

kio_pool_stop(p)
```

Worker death is detected at OS notification latency — a kernel-released
liveness lock is the verdict, with no heartbeats and no polling — and fails
exactly the tasks the dead worker had claimed. If a process crashes without
cleaning up, `kio_prune()` removes the shared-memory regions it left
behind.
