---
title: "Benchmarks"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Two head-to-head measurements against mirai, timed with [bench](https://bench.r-lib.org).
Four mizu workers against four mirai daemons.
Setup is excluded from the timings:

```{r benchmarks-setup}
library(mizu)
library(mirai)

daemons(4L)
p <- mizu_pool(n_workers = 4L)
diamonds <- ggplot2::diamonds

# The task function.
enrich <- evalq(
  function(dat) {
    dat$price_per_carat <- dat$price / dat$carat
    dat
  },
  baseenv()
)

# Speedup multiple of one bench row over another, from the medians
speedup <- function(b, over, base) signif(as.numeric(b$median[over]) / as.numeric(b$median[base]), 2L)
```

**A dataset across the wire, both ways.**
The `enrich()` task receives the 3.3 MB `diamonds` data frame, adds a column, and returns it.
mirai serializes the data once per direction and copies it through a socket.
mizu stages the frame once in shared memory; the worker reads it in place, and the enriched frame crosses back the same way:

```{r benchmark-roundtrip}
roundtrip <- bench::mark(
  mizu = mizu_collect(mizu_submit(p, enrich(dat), dat = diamonds, enrich = enrich)),
  mirai = collect_mirai(mirai(enrich(dat), dat = diamonds, enrich = enrich))
)
roundtrip
```

End to end, mizu is **`r speedup(roundtrip, 2L, 1L)`x** faster: every timed run stages the frame, runs the task — computing the new column over all 53,940 rows — and returns the enriched frame (medians).

**Two thousand small tasks.**
Each task summarizes a sliding window of 1,000 prices — mean and standard deviation, about 10µs of work apiece.
mirai sends one task per element through its dispatcher, whose per-task cost alone exceeds the task.
mizu stages the map once, and the workers self-schedule element ranges off a shared cursor.
The serial `lapply()` row is the baseline a parallel map must beat:

```{r benchmark-map}
winsum <- evalq(function(i) {
  w <- ggplot2::diamonds$price[i:(i + 999L)]
  c(mean = mean(w), sd = stats::sd(w))
}, baseenv())

tasks <- bench::mark(
  serial = lapply(1:2000, winsum),
  mizu = mizu_map(p, 1:2000, winsum),
  mirai = collect_mirai(mirai_map(1:2000, winsum))
)
tasks
```

mizu turns the four workers into a **`r speedup(tasks, 1L, 2L)`x** gain over the serial loop, and runs **`r speedup(tasks, 3L, 2L)`x** ahead of mirai (medians).

```{r benchmarks-teardown}
daemons(0L)
mizu_pool_stop(p)
```

mirai remains the general solution: it scales across machines, where serialization through sockets is unavoidable.
On one machine, mizu removes that cost — an order of magnitude on data movement, and per-task overhead low enough that even microsecond-scale tasks profit from parallelism.

## How mizu compares

mizu occupies a specific point in the design space: parallelism across processes on a single machine, with data moving through shared memory instead of serialized copies.
The tools below overlap at the edges; the differences are in what has to run, how data moves, and where workers can live.

| Tool | What runs | How data moves | Best for |
|------|-----------|----------------|----------|
| `parallel` | Nothing beyond base R | Fork copy-on-write (Unix); socket copies from a cluster | Coarse tasks, maximum portability |
| mirai | Daemon processes | Serialized copies over nanonext sockets | Local or remote workers, HPC clusters |
| crew | A mirai-based, auto-scaled controller | Same as mirai | Bursty workloads and pipelines |
| future | A backend (multisession, multicore, mirai, ...) | Serialized copies; transport per backend | Backend-agnostic parallel loops |
| callr | A fresh R process per call | Serialized copies per call | Isolated one-off computations |
| mizu | Nothing; a package | Shared-memory views, µs handoff | Fine-grained tasks, large vectors and frames, Python interop |

### Base R

`mclapply()` forks workers on Unix — cheap there, because a fork shares the parent's pages copy-on-write — but fork is unavailable on Windows and composes poorly with threads, GUI sessions, and some external libraries; `parLapply()` over a socket cluster is the portable form and serializes every task and result through a socket.

### mirai, crew, and future

mirai dispatches tasks to daemon processes over nanonext sockets, its dispatcher working through callbacks rather than as a broker process — socket copies are the price of its generality: daemons can run anywhere a network reaches, including other machines and HPC clusters, which shared memory cannot.
crew builds auto-scaling worker management on mirai for bursty pipelines.

future layers a backend-agnostic interface over these and other backends (multisession, multicore, or mirai itself via future.mirai), so code ports between them unchanged; the cost of that abstraction is much higher per-task overhead than either mirai or crew, ahead of the backend's own transport.

### callr

callr answers a different question — evaluating R code in a fresh, isolated session with no carry-over of state — and pays for the isolation with a process spawn and a serialized round trip per call.
It is the right tool for one-off isolation, not for dividing a loop.

### The one thing only mizu does

A pool whose workers are Python processes, exchanging vectors and data frames as shared-memory views ([Python interop](interop.html)).
None of the tools above cross the language boundary.

The short version: choose mizu when the work fits one machine, when tasks are fine-grained or vectors and frames large, or when Python needs to be in the loop; choose mirai or crew when workers must run on other machines or under a scheduler, and callr when a task needs a pristine session.
