---
title: "Python interop"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Python interop}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The channel speaks a language-agnostic wire protocol, so the peer at the other end does not have to be R.
[pymizu](https://github.com/shikokuchuo/pymizu) is the Python binding of the same shared-memory core, and a channel connects the two directly.
Pass the peer program as a source string (instead of a quoted expression) and set the launcher to `mizu_py_launcher()`, which spawns `python3 -m pymizu.child` — probing for the interpreter and pymizu before the channel is created.
Here a Python echo loop, mirroring the R one in [Channels](channels.html) (runs when a `python3` with pymizu and NumPy installed is on the `PATH`):

```{r python-interop}
#| eval: !expr nzchar(Sys.which("python3")) && system2("python3", c("-c", shQuote("import pymizu, numpy")), stdout = FALSE, stderr = FALSE) == 0L
library(mizu)

ch <- mizu_channel(
  "
import pymizu
while True:
    x = ch.recv(30)
    if pymizu.is_sentinel(x):
        break
    ch.send(x * 2)
",
  launcher = mizu_py_launcher()
)

mizu_send(ch, c(1.5, 2.5, 3.5))
mizu_recv(ch, timeout = 5)
mizu_close(ch)
```

Atomic vectors cross without serialization: an R `numeric`, `integer`, or `raw` vector arrives in Python as a NumPy `float64`, `int32`, or `uint8` array, and the same dtypes travel back the other way.
An `integer64` vector crosses as a NumPy `int64` array both ways — bit64's exact layout, constructed without bit64 — with `NA` at `INT64_MIN` in each direction.
A large vector arrives as a zero-copy read-only view over the shared pages — the same ALTREP mechanism as between R processes, surfaced on the Python side as a NumPy array that maps the region.
Strings cross as strings in both directions; `NA_character_` reads as `None` in Python, and `None` comes back as `NULL`.

Scalars, lists, and dicts with string keys cross both ways as R's own logical/integer/double/complex/character/list values.
A `Date` arrives as `datetime64[D]`, a `POSIXct` as naive `datetime64[us]`, a factor as `list[str | None]` (a dictionary-encoded column inside a frame), and a matrix as a Fortran-order array — the data never transposes.
A `difftime`, `POSIXlt`, a named vector, and S4 objects stay on R's side: sending one on a Python channel raises at `mizu_send()`.

A data frame crosses as a `pymizu.Frame` — column dict on one side, `pl.from_arrow(f)` / `pa.table(f)` on the way into polars or pyarrow — and a polars, pyarrow, or pandas frame arrives back as a `data.frame`.
Past a size floor the frame travels as one shared-memory region both directions: no copy, no parse, numeric columns read in place.
An unmodified R → polars → R round trip moves zero payload bytes at all — the return hop is a reference to the region R itself staged.
The full contract, with the NA-sentinel and widening rules per dtype, is the dtype matrix in the pymizu README; a value outside the portable subset raises at send time on either side, naming the path and reason.

## Mixed-language pools

Channels move values; pools move computation — a pool's workers can be Python processes, driven through a neutral task specification.
`mizu_py_pool_launcher()` spawns them (`python3 -m pymizu.worker`, probed like the channel launcher), `mizu_call()` describes the task — a qualified `"mod.fn"` name or a `source =` string, plus the constant arguments — and `mizu_submit_call()` stages it:

```{r python-pool}
#| eval: !expr nzchar(Sys.which("python3")) && system2("python3", c("-c", shQuote("import pymizu, numpy")), stdout = FALSE, stderr = FALSE) == 0L
p <- mizu_pool(2L, launcher = mizu_py_pool_launcher())
t <- mizu_submit_call(p, mizu_call("numpy.quantile", runif(100), q = c(0.25, 0.5, 0.75)))
mizu_collect(t)
s <- mizu_submit_call(p, mizu_call(source = "y = x * 2\ny + 1", x = 20))
mizu_collect(s)
mizu_pool_stop(p)
```

Unnamed arguments map to the positional list and named ones to the named dict, matching Python's `*args, **kwargs`; a `source =` task evaluates in a fresh namespace with the named arguments bound as names and the result as the trailing expression's value.
The workers' language comes from the pool itself — a plain `mizu_submit()` on a foreign pool errors locally naming `mizu_submit_call()`, a bare unqualified name errors at submit, and a non-portable argument raises `mizu_error_not_portable` at submit.
Task errors cross as `mizu_error_remote` conditions with the remote type preserved, and a result without a portable home fails the task with one naming the type.
The reverse direction mirrors: `pymizu.Pool.create()` with `pymizu.r_pool_launcher()`, and `Pool.submit()` with `pymizu.call()` specs.

## The cross-language map

A spec is also `mizu_map()`'s `f` — the cross-language map.
The element fills the spec's first positional slot (name kind) or binds as `x` (source kind), the spec's own constants ride with it (so `...` stays empty), and `.template`, `.collect = "view"`, and `.seed` all work — the seed carrying as a neutral pair each worker language derives its own streams from, so invariance holds per worker language rather than identical draws across them:

```{r python-map}
#| eval: !expr nzchar(Sys.which("python3")) && system2("python3", c("-c", shQuote("import pymizu, numpy")), stdout = FALSE, stderr = FALSE) == 0L
p <- mizu_pool(2L, launcher = mizu_py_pool_launcher())
mizu_map(p, runif(6), mizu_call("numpy.quantile", q = c(0.25, 0.75)))
mizu_map(p, runif(6), mizu_call("numpy.log"), .template = numeric(1))
mizu_pool_stop(p)
```
