---
title: "Parallel map"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Parallel map}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`mizu_map()` maps a function over a vector or list on the pool and returns the results in input order.
It is not a loop over `mizu_submit()`.
The function, its constant arguments, and the data are staged once in shared memory.
A handful of chunk tasks divide the elements, and each worker sets up the map at most once.
The per-element cost approaches the cost of `lapply()`, while the work spreads across workers and balances itself through stealing.

```{r map}
library(mizu)

p <- mizu_pool(n_workers = 4L)
mizu_map(p, 1:5, \(i) i * 2L)
```

## Templates

A `.template` (in the style of the `FUN.VALUE` argument of `vapply()`) returns an atomic vector or matrix instead of a list.
The workers write the results straight into shared memory: each result crosses the process boundary exactly once, unserialized:

```{r map-template}
mizu_map(p, seq.int(-5, 5), abs, .template = numeric(1))
```

With `.collect = "view"`, the result is the shared output area itself, wrapped as a read-only ALTREP view: the gather copy is skipped, and the region's teardown defers to the view's garbage collection.

An `integer64` template (bit64's layout, constructed without bit64) is also supported.
int64 joins no widening lattice, so results must be exact `integer64` of the template's length.

## Reproducible randomness

Random numbers drawn inside the function are not reproducible by default, and cost nothing extra.
Pass `.seed` to give every element its own L'Ecuyer-CMRG stream.
Then the results are identical for any chunking, worker count, or steal order:

```{r map-seed}
identical(mizu_map(p, 1:4, \(i) rnorm(i), .seed = 123L),
          mizu_map(p, 1:4, \(i) rnorm(i), .seed = 123L, .chunks = 4L))
```

## Prepared maps

`mizu_map_prepare()` stages a map — the function, its constant arguments, and the data — without running it: the serialization and the region create are paid once.
Each `mizu_map_run()` then costs only task submission and collection, re-arms in O(1), and reuses the cached map context of each worker.
Repeated stochastic simulation is the headline use: `.seed` rides each run rather than the region, so the random streams vary per run for free:

```{r map-prepared}
pm <- mizu_map_prepare(p, 1:1000, \(i, draws) mean(rnorm(draws)) * i, draws = 100L)
runs <- lapply(1:4, \(s) mizu_map_run(pm, .seed = s))
vapply(runs, \(r) r[[1L]], numeric(1))
```

`mizu_map_run(pm, .x = x2)` replaces the staged data: an atomic vector of the same type and length swaps in place at memcpy cost; any other replacement restages transparently.
A run collected with `.collect = "view"` hands its region to the view, so the next run restages into a fresh one.
The handle pins the staged data and the region for its lifetime; both release at garbage collection.

```{r map-stop}
mizu_pool_stop(p)
```
