---
title: "mizu - Reference"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{mizu - Reference}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a reference vignette of the package's core functionality.

mizu provides parallel computation and data exchange between R processes on the same machine: lock-free channels and work-stealing task pools over POSIX shared memory (Linux, macOS) or Win32 file mappings (Windows).
A channel is a two-way message link between an R session and a helper process that it spawns.
A pool is a set of worker processes that divide submitted tasks among themselves.
In both, one process writes data and the other reads it in place — never copied through a socket, pipe, or file.

## Channels

The essential function is `mizu_channel()`.
It creates a two-way shared-memory channel — one lock-free ring per direction — and spawns a child R process connected to the other end.
The child evaluates a quoted expression with `ch` bound to its side of the channel.
All data crosses the rings, not the process boundary.

```{r channel}
library(mizu)

ch <- mizu_channel(quote(
  repeat {
    x <- mizu_recv(ch, timeout = 30)
    if (inherits(x, "mizu_sentinel")) break
    mizu_send(ch, x * 2)
  }
))

mizu_send(ch, 21)
mizu_recv(ch, timeout = 5)
```

`mizu_send()` publishes a message to the peer, visible the moment the call returns.
`mizu_recv()` returns the next message.
`mizu_send_batch()` and `mizu_recv_batch()` move a whole list of messages in one call.
Use them at rates where the per-call overhead of R itself starts to matter.

Outcomes that end a conversation — ring full, timeout, orderly close, peer death — come back as class-tagged sentinel values, not errors.
A receive loop tests for them with `inherits(x, "mizu_sentinel")`, or with the specific classes `mizu_full`, `mizu_timeout`, `mizu_closed`, `mizu_peer_gone`.
No call needs an error handler.
If the process at the other end dies, receives first drain the messages that it already published, then report `mizu_peer_gone`.
`mizu_alive(ch)` reports at any time whether the peer still runs, without touching the rings.

`mizu_close()` performs an orderly shutdown.
It waits for the peer to finish draining before it releases the shared resources:

```{r close}
mizu_close(ch)
```

## Task pools

`mizu_pool()` builds on the same transport.
It spawns a pool of worker processes with work-stealing deques and no dispatcher in the loop.
A submitted task goes straight from the submitting process into shared memory, where a worker claims it.
Idle workers steal from busy ones, so the load balances itself.
`mizu_submit()` captures an expression together with the values it needs and returns a task handle immediately.
Your session stays free to continue.
`mizu_collect()` waits for the result of that task:

```{r pool}
p <- mizu_pool(n_workers = 4L)

t <- mizu_submit(p, sum(x) + y, x = 1:10, y = 100)
mizu_collect(t)
```

If a task raises an error, the pool captures it and signals it again in your session when you collect the result.
`mizu_cancel(t)` withdraws a task.
A task still queued is skipped.
A task already running completes, and its result is discarded.
Cancellation never interrupts executing code.

Inside a task, `mizu_current_pool()` returns the handle of the evaluating worker, so a task can split itself into subtasks.
A nested submit pushes straight onto the own work-stealing deque of the worker — no ring, no wake.
A worker that waits on a nested result executes other work instead of sleeping.
Divide-and-conquer runs at fork/join cost and never deadlocks the pool:

```{r nested}
t <- mizu_submit(
  p,
  {
    subtasks <- lapply(parts, \(part) mizu_submit(mizu_current_pool(), sum(x), x = part))
    do.call(sum, lapply(subtasks, mizu_collect))
  },
  parts = split(1:1000, rep(1:4, each = 250))
)
mizu_collect(t)
```

## Parallel map

`mizu_map()` maps a function over a vector or list on the pool and returns the results in input order.
It is not a loop over `mizu_submit()`.
The function, its constant arguments, and the data are staged once in shared memory.
A handful of chunk tasks divide the elements, and each worker sets up the map at most once.
The per-element cost approaches the cost of `lapply()`, while the work spreads across workers and balances itself through stealing.

```{r map}
mizu_map(p, 1:5, \(i) i * 2L)
```

A `.template` (in the style of the `FUN.VALUE` argument of `vapply()`) returns an atomic vector or matrix instead of a list.
The workers write the results straight into shared memory: each result crosses the process boundary exactly once, unserialized:

```{r map-template}
mizu_map(p, seq.int(-5, 5), abs, .template = numeric(1))
```

An `integer64` template (bit64's layout, constructed without bit64) is also supported.
int64 joins no widening lattice, so results must be exact `integer64` of the template's length.

Random numbers drawn inside the function are not reproducible by default, and cost nothing extra.
Pass `.seed` to give every element its own L'Ecuyer-CMRG stream.
Then the results are identical for any chunking, worker count, or steal order:

```{r map-seed}
identical(mizu_map(p, 1:4, \(i) rnorm(i), .seed = 123L),
          mizu_map(p, 1:4, \(i) rnorm(i), .seed = 123L, .chunks = 4L))
```

## Sizing, sharing and watching a pool

A pool can grow and shrink while it runs.
Retirement is graceful: the worker finishes its current task, and the remaining workers consume anything still queued to it.
Other R processes can join a running pool as submitters — the name of the pool is the only thing they need:

``` r
mizu_spawn_workers(p, n = 2L)     # two more workers join the pool
mizu_retire_worker(p, slot = 0L)  # worker 0 exits after its current task

# in another R process — submit and collect exactly as the creator does:
p2 <- mizu_pool_attach(name)      # name: mizu_pool_status(p)$name on the creator
```

Four read-only tools observe a running pool without disturbing it:

``` r
mizu_pool_status(p)  # snapshot: worker states, queued tasks, result slots
mizu_pool_stats(p)   # cumulative counters: tasks run, steals, parks per worker
mizu_pool_dump(p)    # every slot in full detail — the first tool when a pool hangs
mizu_pool_trace(p, \(event, id) message(event, " ", id))  # task lifecycle hook
```

When you are done, `mizu_pool_stop()` cancels the pending tasks, waits for the workers to exit cleanly, and releases the shared region:

```{r stop}
mizu_pool_stop(p)
```

## Python interop

The channel speaks a language-agnostic wire protocol, so the peer at the other end does not have to be R.
[pymizu](https://github.com/shikokuchuo/pymizu) is the Python binding of the same shared-memory core, and a channel connects the two directly.
Pass the peer program as a source string (instead of a quoted expression) and set the launcher to `mizu_py_launcher()`, which spawns `python3 -m pymizu.child` — probing for the interpreter and pymizu before the channel is created.
Here a Python echo loop, mirroring the R one above (runs when a `python3` with pymizu and NumPy installed is on the `PATH`):

```{r python-interop}
#| eval: !expr nzchar(Sys.which("python3")) && system2("python3", c("-c", shQuote("import pymizu, numpy")), stdout = FALSE, stderr = FALSE) == 0L
ch <- mizu_channel(
  "
import pymizu
while True:
    x = ch.recv(30)
    if pymizu.is_sentinel(x):
        break
    ch.send(x * 2)
",
  launcher = mizu_py_launcher()
)

mizu_send(ch, c(1.5, 2.5, 3.5))
mizu_recv(ch, timeout = 5)
mizu_close(ch)
```

Atomic vectors cross without serialization: an R `numeric`, `integer`, or `raw` vector arrives in Python as a NumPy `float64`, `int32`, or `uint8` array, and the same dtypes travel back the other way.
An `integer64` vector crosses as a NumPy `int64` array both ways — bit64's exact layout, constructed without bit64 — with `NA` at `INT64_MIN` in each direction.
A large vector arrives as a zero-copy read-only view over the shared pages — the same ALTREP mechanism as between R processes, surfaced on the Python side as a NumPy array that maps the region.
Strings cross as strings in both directions; `NA_character_` reads as `None` in Python, and `None` comes back as `NULL`.
Anything else Python sends — scalars and containers ride its compact codec, other objects pickle — is declined with a "Python payload" error.
The declined message is consumed, so the channel keeps flowing; keep the conversation to vectors and strings.

## Benchmarks

Two head-to-head measurements against mirai, timed with [bench](https://bench.r-lib.org).
Four mizu workers against four mirai daemons.
Setup is excluded from the timings:

```{r benchmarks-setup}
library(mirai)

daemons(4L)
p <- mizu_pool(n_workers = 4L)
diamonds <- ggplot2::diamonds

# The task function.
enrich <- evalq(
  function(dat) {
    dat$price_per_carat <- dat$price / dat$carat
    dat
  },
  baseenv()
)

# Speedup multiple of one bench row over another, from the medians
speedup <- function(b, over, base) signif(as.numeric(b$median[over]) / as.numeric(b$median[base]), 2L)
```

**A dataset across the wire, both ways.**
The `enrich()` task receives the 3.3 MB `diamonds` data frame, adds a column, and returns it.
mirai serializes the data once per direction and copies it through a socket.
mizu stages the frame once in shared memory; the worker reads it in place, and the enriched frame crosses back the same way:

```{r benchmark-roundtrip}
roundtrip <- bench::mark(
  mizu = mizu_collect(mizu_submit(p, enrich(dat), dat = diamonds, enrich = enrich)),
  mirai = collect_mirai(mirai(enrich(dat), dat = diamonds, enrich = enrich))
)
roundtrip
```

End to end, mizu is **`r speedup(roundtrip, 2L, 1L)`x** faster: every timed run stages the frame, runs the task — computing the new column over all 53,940 rows — and returns the enriched frame (medians).

**Two thousand small tasks.**
Each task summarizes a sliding window of 1,000 prices — mean and standard deviation, about 10µs of work apiece.
mirai sends one task per element through its dispatcher, whose per-task cost alone exceeds the task.
mizu stages the map once, and the workers self-schedule element ranges off a shared cursor.
The serial `lapply()` row is the baseline a parallel map must beat:

```{r benchmark-map}
winsum <- evalq(function(i) {
  w <- ggplot2::diamonds$price[i:(i + 999L)]
  c(mean = mean(w), sd = stats::sd(w))
}, baseenv())

tasks <- bench::mark(
  serial = lapply(1:2000, winsum),
  mizu = mizu_map(p, 1:2000, winsum),
  mirai = collect_mirai(mirai_map(1:2000, winsum))
)
tasks
```

mizu turns the four workers into a **`r speedup(tasks, 1L, 2L)`x** gain over the serial loop, and runs **`r speedup(tasks, 3L, 2L)`x** ahead of mirai (medians).

```{r benchmarks-teardown}
daemons(0L)
mizu_pool_stop(p)
```

mirai remains the general solution: it scales across machines, where serialization through sockets is unavoidable.
On one machine, mizu removes that cost — an order of magnitude on data movement, and per-task overhead low enough that even microsecond-scale tasks profit from parallelism.

## Deploying on Linux

On Linux, POSIX shared memory is `/dev/shm`, a bounded tmpfs mount.
Everything mizu creates must fit in it: pool and channel regions, plus the per-payload regions that carry anything past the in-slot budget.
Container runtimes set it to 64 MB by default.
Use `docker run --shm-size=4g`, or the equivalent in your orchestrator.
Size it to your peak in-flight payload traffic.

## Linux memory allocator

This section applies only to Linux with glibc.

When mizu starts a channel peer or a pool worker, that process changes two settings of the C memory allocator.
It raises the mmap threshold to 32 MB and the trim threshold to 128 MB.
This keeps large payloads in fast memory.
Without this change, glibc asks the kernel to map and unmap each large payload, and that work is slow.

Your own R process does not change.
If you want the same settings there, set them before R starts:

``` sh
export GLIBC_TUNABLES=glibc.malloc.mmap_threshold=33554432:glibc.malloc.trim_threshold=134217728
```

If you have set `GLIBC_TUNABLES`, mizu respects your values.

## When a process dies

The death of any participant is detected at OS notification latency, with no heartbeats and no polling.
Every process holds a lock that the kernel releases the instant the process exits, for any reason.
That release is the verdict.
A dead worker fails exactly the tasks that it claimed, and collecting one of them raises an error.
The surviving workers consume the work still queued to the dead worker.
On a channel, the survivor sees `mizu_peer_gone`.
This split is deliberate and holds across the whole surface.
Transport states — not yet, not now, stream over — return as sentinel values for the receiving loop to handle.
A request that can never be satisfied raises a classed condition instead (see `?mizu_error`).
Examples: the error of the task itself, a cancelled task, a dead worker, a child that failed to start.

A crashed process cannot clean up after itself.
`mizu_prune()` removes the shared-memory regions that dead processes leave behind.
Regions of running processes are never touched.

## Functions at a glance

| Function | Purpose |
|----------|---------|
| **Channels** | |
| `mizu_channel()` | create a channel and spawn the peer process at the other end |
| `mizu_send()` | send a message to the peer |
| `mizu_recv()` | receive the next message (waits up to a timeout) |
| `mizu_send_batch()` / `mizu_recv_batch()` | move many messages in one call |
| `mizu_alive()` | is the peer process still running? |
| `mizu_close()` | orderly shutdown of a channel |
| **Task pools** | |
| `mizu_pool()` | create a pool and spawn its workers |
| `mizu_submit()` | submit an expression to the pool, get a task handle immediately |
| `mizu_collect()` | wait for and return the result of a task |
| `mizu_cancel()` | withdraw a task (never interrupts one already running) |
| `mizu_map()` | map a function over a vector on the pool, staged once, in input order |
| `mizu_pool_attach()` | join an existing pool as a submitter, from another process |
| `mizu_spawn_workers()` / `mizu_retire_worker()` | grow / shrink the worker set while the pool runs |
| `mizu_pool_status()` / `mizu_pool_stats()` / `mizu_pool_dump()` | observe a pool: snapshot, cumulative counters, full diagnostic dump |
| `mizu_pool_trace()` | register a hook called at each task lifecycle event |
| `mizu_pool_stop()` | shut the pool down |
| **Housekeeping** | |
| `mizu_is_sentinel()` | is this one of the sentinel values of mizu? (identity, not class) |
| `mizu_prune()` | remove shared-memory regions orphaned by crashed processes |
